Metric Spaces with Expensive Distances

### 2018-11-30 paper/draft.tex
Migrate from Overleaf

%\def\marrow{\marginpar[\hfill$\longrightarrow$]{$\longleftarrow$}}
%\def\michael#1{\textcolor{red}{\textsc{Michael says: }{\marrow\sf #1}}}

@@

@@

\subsection{Motivation.}

@@

The next option is to make an additional assumption: while
computing $\dist(\cdot, \cdot)$ exactly is costly, we have access to a cheap approximation algorithm
that gives, say, a $2$-approximation of $\dist$ much faster (the output of this approximation algorithm is guaranteed to
lie in $[\dist(p,q), 2 \dist(p,q)]$). In order to simplify notation, we now assume
that all pairwise distances between our points belong to $[1, 2^t]$ for some $t$.
We now introduce \textit{buckets} of the form $[2^k, 2^{k+2}]$ for $k = 0 \dots t$;
these intervals are not disjoint, but, by calling our algorithm we can assign
each pair $(p_i, p_j)$ to one of the buckets.
Indeed, let $v$ be the output of the approximation algorithm for $\dist(p,q)$,
then we know that the true distance is in $[v/2, v]$. We want to put the pair $(p,q)$ into the bucket
that fully contains this interval, thus we need to find $k$ such that
$2^k \leq v /2$ and $v \leq 2^{k+2}$. These inequalities give $ 2^{k+1} \leq v \leq 2^{k+2}$,
and we take $k = \lfloor \log v \rfloor - 1$.
If we find a bucket for each pair $(p_i, p_j)$, we have some sort of a weak ordering: we cannot say anything
about two edges that are assigned to the same bucket or to overlapping buckets, but if the buckets
are disjoint, we know which edge is longer. Now we can either imitate the greedy spanner (non-blind) algorithm,
traversing buckets in the order of their left endpoint, and adding edges from them; or we could alternate: in even iterations
we travers the buckets from left to right, adding edges from each bucket one by one, and in odd iterations
we traverse the buckets from right to left. This alternative makes sense, because we need to have good lower bounds in
order to build a blind spanner, and good lower bounds can be obtained from the triangle inequality, if one of the edges 
in a path connecting $p_i$ and $p_j$ in $G$ is longer than the sum of the lengths of all other edges.
We refer to the former option as \textsc{BlindQuasiSortedGreedy} and to the latter one as \textsc{BlindQuasiSortedShaker}.

@@

Note that after we added the edge $(p_i, p_j)$, the ratio of the upper and lower bound
for this pair becomes 1, so a blind spanner algorithm always terminates (of course,
we assume that the procedure \textsc{GetEdgeToAdd} picks an edge only once).

>>>>

\def\marrow{\marginpar[\hfill$\longrightarrow$]{$\longleftarrow$}}
\def\michael#1{\textcolor{red}{\textsc{Michael says: }{\marrow\sf #1}}}

@@

\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
\subjclass{dummy}
% \category{dummy}

@@

\begin{description}
\item[Approximate Nearest Neighbor] Given $\eps>0$ and a point $q\in\metricspace$,
find $p_i\in\pointset$ such that, for all $j=1,\ldots,n$,
\[\dist(q,p_i)\leq(1+\eps)\dist(q,p_j)\]

\item[Spanner] Given $\eps>0$, compute a weighted graph $G$ with vertices in $\pointset$
such that for any $u,v\in\pointset$, the shortest path distance between $u$ and $v$
is at most $(1+\eps)\dist(u,v)$.
\end{description}

The performance of algorithms for these problems depends on the number of points,
the dimension of the metric space, and the cost $\complexity$ of computing a distance in the metric space.
It is a common assumption to assume $\complexity$ to be a constant; 
There are good reasons for that: the most common case of a metric space
is $\metricspace=\R^d$ with $d$ some constant, in which case $\metric$ can be evaluated in $O(d)=O(1)$ time.
Even if $d$ is considered non-constant,
it can always be assumed that $d\leq n$, hence $\complexity$ is at most $O(n)$.
Another typical assumption is that all pairwise distances are part of the input
in which case $\complexity$ is $O(1)$.

However, we argue that in some situations, distance computations
in $\metricspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivation comes from topological summaries
such as persistence diagrams or Reeb graphs, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\R^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered. For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed,
not even an constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.
\michael{This paragraph needs references. As you know the literature better than me,
I leave that part to you.}

In such situations with expensive distance computations, 
it makes sense to study a different cost model, where only the number of distance computations
is taken into account. For instance, that means that quadratic time operations in terms of $n$
are not counted towards the time complexity, as long as these operations do not query any distance
in $\metricspace$. We also ignore the space complexity in our model.

We will restrict to the case of \emph{doubling spaces}, that is, the doubling dimension
of $\metricspace$ is bounded by a constant. 
In that situation, standard constructions from computational geometry provide partial answers:
Using net-trees~\cite{hm-fast}, we can construct a $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
an $\eps$-spanner immediately. Net-trees can also be used to compute approximate nearest neighbors
performing $O(\log n)$ distance computations per query point.
However, for our relaxed cost model, we pose the question whether simpler constructions achieve
comparable, or even fewer distance computations.

We also propose a slight variant of our model: we assume that we also have access to an (efficient)
$2$-approximation algorithm for the distance queries. Queries to this approximation algorithm
are not counted in the model, hence we can assume that for each pair of points $(u,v)$, we
know a number $a_{u,v}$ with $\dist(u,v)\leq a_{u,v}\leq 2\dist(u,v)$. This induces an approximate ordering
of all distances in the metric space, and it is plausible to assume that such an ordering will simplify
algorithmic tasks on metric spaces, at least in practice.

\myparagraph{Contributions}
%
We propose simple algorithms for spanner construction and approximate nearest neighbor search
and evaluate them theoretically and experimentally in the defined cost model.

Our algorithms are based on the following simple idea: since distance computations are expensive
and should be avoided, we try to obtain maximal information out of the distances that have been computed
so far. 
This information consists of lower and upper bounds for unknown distances, obtained from known distances
by triangle inequality (see Figure~\ref{fig:1st_example}).

\begin{figure}[h]
\centering
\includegraphics[width=6cm]{intro_example.eps}
\caption{The compute distances are shown as edges in a graph. Note that the exact distance
of $p_1$ and $p_2$ is unknown. The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
constitutes an upper bound on the distance by triangle inequality.
However, we can also infer that $\metric(p_1,p_2)\geq 3$:
otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
would be shorter than the edge $(p_3,p_4)$, again contradicting
triangle inequality.}
\label{fig:1st_example}
\end{figure}

we see a weighted graph on $P$ whose edges are the distances that have
been computed so far; note that $\metric(p_1,p_2)$ is unknown. 
The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
constitutes an upper bound on the distance by triangle inequality.
However, we can also infer that $\metric(p_1,p_2)\geq 3$:
otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
would be shorter than the edge $(p_3,p_4)$, again contradicting
triangle inequality.
Hence, the question whether $\metric(p_1,p_2)\leq t$ can be answered positively
for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$
We describe our idea with an example. In Figure~\ref{fig:1st_example},
we see a situation on $P$ whose edges are the distances that have
been computed so far; note that $\metric(p_1,p_2)$ is unknown. 
The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
constitutes an upper bound on the distance by triangle inequality.
However, we can also infer that $\metric(p_1,p_2)\geq 3$:
otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
would be shorter than the edge $(p_3,p_4)$, again contradicting
triangle inequality.
Hence, the question whether $\metric(p_1,p_2)\leq t$ can be answered positively
for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$

@@

The last option that we consider is based on an additional assumption: while
computing $\dist(\cdot, \cdot)$ exactly is costly, we have access to a
cheap approximation algorithm
that gives a $2$-approximation of $\dist$ much faster 
(the output of this approximation algorithm is guaranteed to
lie in $[\dist(p,q), 2 \dist(p,q)]$). In order to simplify notation, we now assume
that all pairwise distances between our points belong to $[1, 2^t]$ for some $t$.
We introduce \textit{buckets} of the form $[2^k, 2^{k+2}]$ for $k = 0 \dots t$;
these intervals are not disjoint, but by calling the approximation algorithm we can assign
each pair $(p_i, p_j)$ to one of the buckets.
Indeed, let $v$ be the output of the approximation algorithm for $\dist(p,q)$,
then we know that the true distance is in $[v/2, v]$. We want to put the pair $(p,q)$ into the bucket
that fully contains this interval, thus we need to find $k$ such that
$2^k \leq v /2$ and $v \leq 2^{k+2}$. These inequalities give $ 2^{k+1} \leq v \leq 2^{k+2}$,
and we take $k = \lfloor \log v \rfloor - 1$.
Afer we found a bucket for each pair $(p_i, p_j)$, 
we have some sort of a weak ordering: we cannot say anything
about two edges that are assigned to the same bucket or to overlapping buckets, but if the buckets
are disjoint, we know which edge is longer. Now we can either imitate the greedy spanner (non-blind) algorithm,
traversing buckets in the order of their left endpoint, and adding edges from them; or we could alternate: in even iterations
we traverse the buckets from left to right, adding edges from each bucket one by one, and in odd iterations
we traverse the buckets from right to left. This alternative makes sense, because we need to have good lower bounds in
order to build a blind spanner, and good lower bounds can be obtained from the triangle inequality, if one of the edges 
in a path connecting $p_i$ and $p_j$ in $G$ is longer than the sum of the lengths of all other edges.
We refer to the former option as \textsc{BlindQuasiSortedGreedy} and to the latter one as \textsc{BlindQuasiSortedShaker}.

@@

% Note that after we added the edge $(p_i, p_j)$, the ratio of the upper and lower bound
% for this pair becomes 1, so a blind spanner algorithm always terminates (of course,
% we assume that the procedure \textsc{GetEdgeToAdd} picks an edge only once).
